Collaboratively Testing the Validity of Neuroscientific Models

### 2014-03-27 neuronunit_frontiers/Paper.tex
Revisions for Frontiers

Software engineers seeking to verify, validate and contribute to a complex software project rely not only on volumes of human documentation, but on suites of simple executable tests, called ``unit tests''. 
Drawing inspiration from this practice, we have developed \textit{SciUnit}, an easy-to-use framework for developing ``model validation tests'' -- 
executable functions, here written in Python. 
These tests generate and statistically validate predictions from a specified class of scientific models against one relevant empirical observation to produce a score indicating agreement between the model and the data. 
Suites of such validation tests, collaboratively developed by a scientific community in common repositories, produce up-to-date statistical summaries of the state of the field. In this paper, we aim to detail this test-driven workflow and introduce it to the neuroscience community. 
As an initial example, we describe \textit{NeuronUnit}, a library that builds upon \textit{SciUnit} and integrates with several existing neuroinformatics resources to support validating single-neuron models using data gathered by neurophysiologists.

@@

Today, scientists contribute a new model to the research community by submitting a paper containing a description of the model's structure and scope along with text and figures that demonstrate its validity and argue for its novelty.  
The scientists tasked with reviewing the paper are  responsible for evaluating these claims, discovering competing models and relevant data the paper did not adequately consider, and ensuring that goodness-of-fit was measured in a statistically sound manner, drawing on their knowledge of prior literature. 
Once a modeling paper is published, it is frozen and cited as-is in perpetuity.

@@

\subsection{Test Suites and Score Matrices}\label{sec:visualization} A collection of tests intended to be run on the same model can be put together to form a test suite.
The following test suite could be used for a simplified version of the QSNMC, as described in sec. \ref{sec:example}:  
\begin{ipy}
  In [5]: CA1_suite = sciunit.TestSuite([CA1_overall_test, CA1_sc_test, CA1_sl_test, CA1_sv_test, 
              CA1_isi_test, CA1_imi_test])
\end{ipy}
Like a single test, a test suite is capable of judging one or more models. 
The models must possess the union of the capabilities of the tests in a suite to be fully capable of running the suite, though partial results can also be generated. 
When a model cannot take a test because of a missing capability, the \texttt{NotApplicable} score is used (e.g. point processes cannot predict subthreshold voltages). 
The result is a \emph{score matrix}, like the one diagrammed in Fig. \ref{fig:sciunit_overview}.
\begin{ipy}
  In [6]: scores = CA1_suite.judge([ARX_model, AdEx1_model, aSRM_model, PP_model])
\end{ipy}
A simple summary of the scores in a score matrix can be printed to the console or visualized by other tools.  
For example, the score matrix can be visualized as a hyperlinked table inside an IPython notebook \citep{perez_ipython_2007} (Fig. \ref{fig:scidash_matrix}).

@@

The most visible result of these efforts will be tables, like the one shown in Figure \ref{fig:scidash_matrix}, that allow scientists to understand, precisely, the state-of-the-art in their field. 
We are developing a lightweight portal, \textit{SciDash}, that directs scientists to the canonical test repositories in their field in order to prevent fragmentation and encourage community participation \citep{omar_sciunit_2013}. 
\textit{SciDash} will also facilitate viewing test results without needing to clone test repositories and run tests locally for exploratory purposes, based on the \textit{iPython notebook} technology mentioned in Sec. \ref{sec:visualization}. 
We leave details of \textit{SciDash} as future work, focusing this paper on the core testing methodology (Sec. 2) and on the details of the bridge technologies, below.

@@

\end{figure}
The NeuroElectro project (\url{http://neuroelectro.org}) is an effort to make all published data on single cell neurophysiology available in a machine-readable format \citep{tripathy_neuroelectro:_2012}.  
Currently, up to 27 electrophysiological properties are available for 93 cell types, gathered from over 2000 single pieces of published data extracted from tables in published papers. 
NeuroElectro uses the NeuroLex.org ontology \cite{larson_neurolex_2013} to identify specific neuron types and specific electrophysiological properties, and indexes corresponding data values from the literature.  
We have made it easy to parameterize the test families in \textit{Neuron\-Unit}  using data extracted from the NeuroElectro API. 
Tests can be based upon data from single journal articles, or from ensembles of articles with a common theme (e.g. about a particular neuron type). 
The latter is illustrated in Figure \ref{fig:neuronunit_example}, on lines 1-4. 
Once data has been retrieved from NeuroElectro, associated statistics (e.g. mean, standard error and sample size) can be extracted to parameterize tests (e.g. a test of spike widths on line 10). 
Data from NeuroElectro can be used to validate a variety of basic electrophysiological features of neuron models, including spike width, resting membrane potential, after-hyperpolarization amplitude, etc. 
As NeuroElectro is the only public, curated source of such data, it represents a key resource facilitating test construction.

@@

\subsection{Participation from Modeling Communities}
Modelers may not want to formally expose their models to tests via capabilities, a requirement for test-taking.  
We anticipate four solutions: \textbf{First}, we will emphasize to the community that wrapping existing model code that satisfy a capability is quite simple, requiring as little as one line of code. 
Importantly, the modeler is not required to expose or rewrite any internal model flow control in almost all cases. 
\textbf{Second}, we support multiple simulation environments automatically by using NeuroML \citep{gleeson_neuroml:_2010}, and other simulator-independent model descriptions are possible for other domains. 
Automated generation of NeuroML from native model source code is in development (Gleeson, personal communication); for the popular NEURON simulator (\cite{carnevale_neuron_2006}), this functionality is already implemented and in use. 
This minimizes modeler effort for a large (at least 1000, \url{http://www.neuron.yale.edu/neuron/node/69}) and growing number of neuronal models. 
\textbf{Third}, modelers have an incentive to demonstrate publicly their models' validity. 
Participation in public modeling competitions demonstrates that this incentive can draw several participants. We plan to assist with such competitions by advocating for use of SciUnit as the underlying technical substrate in the future. 
\textbf{Fourth}, modelers have an incentive to use \textit{SciUnit} during development (see TDD, Sec. \ref{sec:introduction}) to ensure that ongoing development preserves correspondence between model and data. 
A popular test suite can represent a ``gold standard'' by which progress during development is judged, even within a single project.

@@

\subsection{Occam's Razor}
All things being equal, simpler models are better. 
Model complexity, though fundamentally a subjective measure, has many correlates \citep{mccabe_complexity_1976}, including: 
1) model length; 2) memory use; 3) CPU load; 4) \# of capabilities. 
Future \textit{SciUnit} tools will analyze these factors, and may include tools for gathering community ratings about subjective factors like complexity. As a result of these enhancements, scientists will be able to visualize the model validity vs complexity tradeoff in tabular form (e.g. as a column in Figure \ref{fig:scidash_matrix}), or as a scatter plot, with the ``best'' models being in the high validity / low complexity corner of the plot. 
The set of models which \textit{dominate} all others, i.e. that have the highest validity for a given complexity, can be represented as a ``frontier'' in such a scatter plot (e.g. as used in the symbolic regression package Eureqa \citep{schmidt_distilling_2009}).

@@

>>>>

Software engineers seeking to verify, validate and contribute to a complex software project rely not only on volumes of human documentation, but on suites of simple executable tests, called ``unit tests''. 
Drawing inspiration from this practice, we have developed \textit{SciUnit}, an easy-to-use framework for developing ``model validation tests'' -- 
executable functions, here written in Python. 
These tests generate and statistically validate predictions from a specified class of scientific models against one relevant empirical observation to produce a score indicating agreement between the model and the data. 
Suites of such validation tests, collaboratively developed by a scientific community in common repositories, can produce up-to-date statistical summaries of the state of the field. In this paper, we aim to detail this test-driven workflow and introduce it to the neuroscience community. 
As an initial example, we describe \textit{NeuronUnit}, a library that builds upon \textit{SciUnit} and integrates with several existing neuroinformatics resources to support validating single-neuron models using data gathered by neurophysiologists.

@@

Today, scientists contribute a new model to the research community by submitting a paper containing a description of the model's structure and scope along with text and figures that demonstrate its validity and argue for its novelty.  
The scientists tasked with reviewing the paper are responsible for evaluating these claims, discovering competing models and relevant data the paper did not adequately consider, and ensuring that goodness-of-fit was measured in a statistically sound manner, drawing on their knowledge of prior literature. 
Often, there are no means for verifying even the basic results \citep{Donoho0815years}.
And once a modeling paper is published, it is frozen and cited as-is in perpetuity.

@@

\subsection{Test Suites and Score Matrices}\label{sec:visualization} A collection of tests intended to be run on the same model can be put together to form a test suite.
The following test suite could be used for a simplified version of the QSNMC, as described in sec. \ref{sec:example}:  
\begin{ipy}
  In [5]: CA1_suite = sciunit.TestSuite([CA1_overall_test, CA1_sc_test, CA1_sl_test, CA1_sv_test, 
              CA1_isi_test, CA1_imi_test])
\end{ipy}
Like a single test, a test suite is capable of judging one or more models. 
The models must possess the union of the capabilities of the tests in a suite to be fully capable of running the suite, though partial results can also be generated. 
When a model cannot take a test because of a missing capability, the \texttt{NotApplicable} score is used (e.g. point processes cannot predict subthreshold voltages). 
The result is a \emph{score matrix}, like the one diagrammed in Fig. \ref{fig:sciunit_overview}.
\begin{ipy}
  In [6]: scores = CA1_suite.judge([ARX_model, AdEx1_model, aSRM_model, PP_model])
\end{ipy}
A simple summary of the scores in a score matrix can be printed to the console or visualized by other tools.  
For example, the score matrix can be visualized as a hyperlinked table inside an IPython notebook \citep{perez_ipython_2007} (Fig. \ref{fig:scidash_matrix}).  The code for generating such visualizations is available in the main SciUnit repository (\url{http://github.com/scidash/sciunit}), and for generating this particular visualization is available at \url{Cyrus put a URL here}.

@@

The most visible result of these efforts will be tables, like the one shown in Figure \ref{fig:scidash_matrix}, that allow scientists to represent, formally, the state of model development in their field, as judged against corresponding experimental data. 
We are developing a lightweight portal, \textit{SciDash}, that directs scientists to the canonical test repositories in their field in order to prevent fragmentation and encourage community participation \citep{omar_sciunit_2013}. 
\textit{SciDash} will also facilitate viewing test results without needing to clone test repositories and run tests locally for exploratory purposes, based on the \textit{iPython notebook} technology mentioned in Sec. \ref{sec:visualization}. 
We leave details of \textit{SciDash} as future work, focusing this paper on the core testing methodology (Sec. 2) and on the details of the bridge technologies, below.

@@

\end{figure}
The NeuroElectro project (\url{http://neuroelectro.org}) is an effort to make all published data on single cell neurophysiology available in a machine-readable format \citep{tripathy_neuroelectro:_2012}.  
Currently, up to 27 electrophysiological properties are available for 93 cell types, gathered from over 2000 single pieces of published data extracted from tables in published papers. 
NeuroElectro uses the NeuroLex.org ontology \cite{larson_neurolex_2013} to identify specific neuron types and specific electrophysiological properties, and indexes corresponding data values from the literature.  
We have made it easy to parameterize the test families in \textit{Neuron\-Unit}  using data extracted from the NeuroElectro API. 
Tests can be based upon data from single journal articles, or from ensembles of articles with a common theme (e.g. about a particular neuron type). 
The latter is illustrated in Figure \ref{fig:neuronunit_example}, on lines 1-4. 
Once data has been retrieved from NeuroElectro, associated statistics (e.g. mean, standard error and sample size) can be extracted to parameterize tests (e.g. a test of spike widths on line 10). 
Data from NeuroElectro can be used to validate a variety of basic electrophysiological features of neuron models, including spike width, resting membrane potential, after-hyperpolarization amplitude, etc. 
As NeuroElectro is the only public, curated source of such data, it represents a key resource facilitating test construction.  
However, in selecting data from NeuroElectro (either averaged data across all reports, or data from selected reports) for test construction, one is implicitly assuming that it represents the kind data that ought to be binding in model development.  
For some models, data obtained in this way is ought not to guide model development, and other data sources will need to be identified and used for test construction.

@@

\subsection{Participation from Modeling Communities}
Modelers may not want to formally expose their models to tests via capabilities, a requirement for test-taking.  
We anticipate four solutions: \textbf{First}, we will emphasize to the community that wrapping existing model code that satisfy a capability is quite simple, requiring as little as one line of code. 
Importantly, the modeler is not required to expose or rewrite any internal model flow control in almost all cases. 
\textbf{Second}, we support multiple simulation environments automatically by using NeuroML \citep{gleeson_neuroml:_2010}, and other simulator-independent model descriptions are possible for other domains. 
Automated generation of NeuroML from native model source code is in development (Gleeson, personal communication); for the popular NEURON simulator (\cite{carnevale_neuron_2006}), this functionality is already implemented and in use. 
This minimizes modeler effort for a large (at least 1000, \url{http://www.neuron.yale.edu/neuron/node/69}) and growing number of neuronal models. 
\textbf{Third}, modelers have an incentive to demonstrate publicly their models' validity. 
Participation in public modeling competitions demonstrates that this incentive can draw several participants. We plan to assist with such competitions by advocating for use of SciUnit as the underlying technical substrate in the future. 
Currently, we are in discussion with field specialists to organize competitions for spiking neuron models (a sequel to the QSNMC), spike-sorting algorithms, calcium imaging spike time reconstruction, brain-computer interfaces, neural network function prediction, and seizure prediction.  
\textbf{Fourth}, modelers have an incentive to use \textit{SciUnit} during development (see TDD, Sec. \ref{sec:introduction}) to ensure that ongoing development preserves correspondence between model and data. 
A popular test suite can represent a ``gold standard'' by which progress during development is judged, even within a single project.

@@

\subsection{Occam's Razor}
All things being equal, simpler models are better. 
Model complexity, though fundamentally a subjective measure, has many correlates \citep{mccabe_complexity_1976}, including: 
1) model length; 2) memory use; 3) CPU load; 4) \# of capabilities. 
Future \textit{SciUnit} tools will analyze these factors, and may include tools for gathering community ratings about subjective factors like complexity. As a result of these enhancements, scientists will be able to visualize the model validity vs complexity tradeoff in tabular form (e.g. as a column in Figure \ref{fig:scidash_matrix}), or as a scatter plot, with the ``best'' models being in the high validity / low complexity corner of the plot. 
Validity may be equated to cross-validation accuracy, or the inverse of some loss function.
The set of models which \textit{dominate} all others, i.e. that have the highest validity for a given complexity, can be represented as a ``frontier'' in such a scatter plot (e.g. as used in the symbolic regression package Eureqa \citep{schmidt_distilling_2009}).

@@

\subsection{Project Development}
Gewaltig and Cannon describe the maturity of a computational software project \citep{gewaltig_current_2014} as ``review ready'', ``research ready'', or ``user ready''.  
The development of the underlying SciUnit framework is complete \citep{omar_sciunit_2014}, and it provides the full interface necessary for continued development of domain-specific unit-testing frameworks -- it is research ready.  NeuronUnit is ``review ready'' -- we invite others to participate in building it out to satisfy the use cases they encounter.  Corresponding test suite repositories on GitHub, containing NeuronUnit-based tests, do not even reach this stage.  With the exception of the repositories in development for the QSNMC, and for the Open Source Brain pipeline describe in Section 4, they are non-existent.  We plan to continue creating such repositories; more importantly we invite interested parties to create their own, either to guide in-house model development or to assess larger pools of established models.
